Towards a Neural Co-Processor Which Restores Movement After Stroke: Modeling a Proof-of-Concept

### 2021-11-14 paper/article.tex
Paper progress

Before attempting \textit{in vivo} experiments using such a model, we first demonstrate
here a number of crucial design elements of it, through the use of a simulated
grasping circuit, presented previously by Michaels et al. \cite{michaels.mrnn}. We
explore:
\begin{itemize}
	\item The properties of the artificial neural networks (ANNs) that are needed to
	      successfully adapt to the long-running dynamics of a stimulated neural circuit,
	      as well as to adapt to that circuit's ongoing changes.
	\item Training methods for our stimulation model, to make it effective in later
	      training our stimulation agent.
	\item Data reuse methods, to improve learning efficiency, to better ensure we
	      can train them with a biologically-realistic amount of data.
\end{itemize}

@@

Note that the EN effectively constitutes a stimulation model. We initially attempted
an EN architecture resembling the common linear time-invariant state space model of
stimulation, as in e.g. Yang et al. \cite{shanechi.stimmodel}, which is a helpfully
simple (linear) approach. That approach seeks to model neural network dynamics in
terms of a stationary linear model, which, due in part to its simplicity and the
number of techniques developed for it, lends itself to useful interpretations and
analyses of those neural dynamics. We found, however, that the linear approach,
as well as the related vanilla RNN approach with nonlinearities added, was not
sufficiently expressive of long-term dependencies, such that the co-processor
was capable of learning well.

Separately, nonlinear LSTM models have also been applied to modeling natural
neural networks, such as use in predicting local field potentials
\cite{kim.lstm}, and for stimulation modeling \cite{guclu.lstm}.
LSTM cells are designed to better capture the relevant aspects of
long-running network dynamics, and in our case we found that they were crucial
for that reason. As a result, both our EN and CPN models leverage LSTM cells.

@@

Clearly, training and deploying a neural co-processor in a true brain involves challenges we
don't capture with this simulation. We don't intend this simulation to constitute evidence
that our method will, for example, immediately translate to true restoration of fine-grained
control for individual fingers of a stroke patient. One clear challenge with doing so would be the
sheer dimensionality of natural brain dynamics, and how that compares to the resolution
of stimulation the co-processor can learn to apply. There exists a clear mismatch between
the dimensionality of the underlying problem, electrode technologies, and the amount of training
data which can reasonably be collected to train a closed-loop neural controller. We argue
that while this fact clearly creates a challenge for learning-based closed-loop stimulation,
that nevertheless the insights we are generating through this simulation are likely to be
applicable. As we show next, the simulation exhibits a number of properties that reflect
a real world application, which our co-processor design must contend with. In the Discussion
section below, we will explore how those insights are likely to lead to initial demonstrations
of the co-processor approach on lower dimensional problems.

@@

\subsubsection{Simulating recovery prior to co-processor use}
We sought to additonally pre-train the mRNN after applying the lesion in
order to simulate stroke recovery. For simulated lesions which zero the outputs
of neurons, the mRNN is, unfortunately, not an appropriate model.
We know that strokes can cause damage from which the subject, in many cases,
cannot recover. The mRNN model has sufficient redundancy built into it that
lesioning the outputs of large numbers of neurons leaves enough remaining
degrees of freedom that a nearly full recovery can occur, unless so many
neurons are lesioned that no stimulation could be effective. We go into
this further in the Discussion section below.

@@

Addiontally, we use PyTorch's $AdamW$ optimizer, which includes an option for weight regularization.
With that, we use a carefully chosen learning rate schedule. Our intuition is to leverage regularization as
one of the standard mitigations for overfit. To converge, the learning rate schedule needs
to be chosen carefully: a relatively high learning rate at first, dropping to a very low rate soon
thereafter. Using too-low or too-high of a learning rate in any phase causes the EN learning
to not converge. EN training proceeds until it reaches a threshold of prediction error, defined as
a fraction of the current CPN's task loss.

@@

\section{Discussion and Conclusion}
\begin{itemize}
	\item Co-processors can be trained, and they can adapt to a difficult problem
	\item One central issue that still needs work is data/training efficiency
	\item Pitch a couple ideas to help solve it:
		\begin{itemize}
			\item Continuous retraining of EN, rather than throwing it away
			\item Retain recent data and reuse it during training
		\end{itemize}
	\item Toward in vivo application: example of lower dimensional problems this may help with
\end{itemize}

@@

\subsection{Future application}
(on a lower dimensional problem presumably)

@@

\section{Ethical Statement}
TODO.

>>>>

Before attempting \textit{in vivo} experiments using such a model, we first demonstrate
here a number of crucial design elements of it, through the use of a simulated
grasping circuit, presented previously by Michaels et al. \cite{michaels.mrnn}. We
explore the properties of the artificial neural networks (ANNs) that are needed to
successfully adapt to the long-running dynamics of a stimulated neural circuit,
as well as to adapt to that circuit's ongoing connectivity changes.
We additionally present a training method our co-processor, allowing it to learn
an optimal stimulation pattern that drives improvements in external task performance,
while also adapting to the non-stationarity of the brain.

@@

Note that the EN effectively constitutes a stimulation model. However, in the case that
a co-processor attempts to optimize for task performance, as opposed to neural
activity directly, the stimulation model predicts the effect of stimulation and neural
dynamics on task performance, rather than modeling and predicting the neural response
to stimulation alone. This provides the key functionality needed to train the CPN.

We initially attempted an EN architecture using vanilla RNN cells, with a nonlineaerity.
This approach resembles the common linear time-invariant state space model of
stimulation, as in e.g. Yang et al. \cite{shanechi.stimmodel}, though it additionally
incorporates a nonlinearity. That approach has the benefit of being simple, and familiar.
We found, however, that the linear approach, as well as the related vanilla RNN
approach with nonlinearities added, was not sufficiently expressive of long-term
dependencies, such that the co-processor was capable of learning well.

Separately, nonlinear LSTM models have also been applied to modeling natural
neural networks, such as use in predicting local field potentials
\cite{kim.lstm}, and for stimulation modeling \cite{guclu.lstm}.
LSTM cells are designed to better capture the relevant aspects of
long-running network dynamics. We found that LSTM cells were crucial for
enabling our co-processor to learn the long-running dynamics of the network
it was stimulating. As a result, both our EN and CPN models leverage LSTM cells.

@@

Clearly, training and deploying a neural co-processor in a true brain involves challenges we
don't capture with this simulation. We don't intend this simulation to constitute evidence
that our method will, for example, immediately translate to true restoration of fine-grained
control for individual fingers of a stroke patient. One clear challenge with doing so would be the
sheer dimensionality of natural brain dynamics, and how that compares to the resolution
of stimulation the co-processor can learn to apply. There exists a clear mismatch between
the dimensionality of the underlying problem, sensor and stimulator technologies, and the amount of training
data which can reasonably be collected to train a closed-loop neural controller. We argue
that while this fact clearly creates a challenge for learning-based closed-loop stimulation,
that nevertheless the insights we are generating through this simulation are likely to be
applicable. In the discussion below, we will explore how those insights are likely to lead
to initial demonstrations of the co-processor approach on lower dimensional problems.
As we show next, the simulation exhibits a number of properties that reflect
a real world application, which our co-processor design must contend with.

@@

\subsubsection{Simulating recovery prior to co-processor use}
We sought to additonally re-train the mRNN after applying the lesion in
order to simulate stroke recovery. For simulated lesions which zero the outputs
of neurons, the mRNN is, unfortunately, not an appropriate model.
We know that strokes can cause damage from which the subject, in many cases,
cannot recover. The mRNN model has sufficient redundancy built into it that
lesioning the outputs of large numbers of neurons leaves enough remaining
degrees of freedom that a nearly full recovery can occur, unless so many
neurons are lesioned that no stimulation could be effective.

@@

Addtionaly, we use PyTorch's $AdamW$ optimizer, which includes an option for weight regularization.
With that, we use a carefully chosen learning rate schedule. Our intuition is to leverage regularization as
one of the standard mitigations for overfit. To converge, the learning rate schedule needs
to be chosen carefully: a relatively high learning rate at first, dropping to a very low rate soon
thereafter. Using too-low or too-high of a learning rate in any phase causes the EN learning
to not converge. EN training proceeds until it reaches a threshold of prediction error, defined as
a fraction of the current CPN's task loss.

@@

\section{Discussion}
\label{sec:discussion}

We present here a first-of-its-kind demonstration of a novel design for neural co-processors,
allowing for an AI agent to learn stimulation policies that improve a user's task performance.
The central notion of the design is to leverage a stimulation model, whose purpose is to provide
training feedback to an AI agent responsible for generating the stimulation. We present here
a first demonstration of the learning algorithms and key properties

We proposed X
It did Y
It was found that Z
It has potential to A

\begin{itemize}
	\item Co-processors can be trained, and they can adapt to a difficult problem
	\item One central issue that still needs work is data/training efficiency
	\item Pitch a couple ideas to help solve it:
		\begin{itemize}
			\item Continuous retraining of EN, rather than throwing it away
			\item Retain recent data and reuse it during training
		\end{itemize}
	\item Toward in vivo application: example of lower dimensional problems this may help with
\end{itemize}

@@

\section{Conclusion}
...
\subsection{Future application}
(on a lower dimensional problem presumably)

@@
